---
title: "Ira_Tana_DGE_GSEA"
author: "IH_TS"
date: "2025-04-18"
output:
  html_document:
    code_folding: hide
    df_print: paged
    highlights: pygments
    number_sections: yes
    self_contained: yes
    theme: default
    toc: yes
    toc_float:
      collapsed: yes
  pdf_document: 
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
##Install packages
```{r echo=TRUE, results="hide", message=FALSE, warning=FALSE}
install.packages('remotes')
remotes::install_github("satijalab/seurat", "seurat5", quiet = TRUE)
install.packages("renv")
install.packages("Seurat")
install.packages("ggplot2")
install.packages("ggrepel")
install.packages("enrichR")

```
##Import packages
```{r echo=TRUE, results="hide", message=FALSE, warning=FALSE}
library(dplyr)
library(magrittr)
library(Seurat)
library(ggplot2)
library(ggrepel)
library(enrichR)
```

## Read seurat_object_after_clustering

```{r, echo=TRUE, results="hide", message=FALSE, warning=FALSE}
seurat_obj <- readRDS("C:/Users/Doma/Desktop/RSTUDIO/ProjectAML - kopie - kopie/seurat_object_after_clustering.rds")
seurat_obj
table(Idents(seurat_obj))
View(seurat_obj@meta.data)
```
##DGE

## Processing HSC cluster and find markers AML vs Control
```{r, echo=TRUE, results="hide", message=FALSE, warning=FALSE}
subset_cluster_HSC <- subset(seurat_obj, idents = "HSC")
table(subset_cluster_HSC@meta.data$group)
Idents(subset_cluster_HSC) <- "group"


#Find markers HSC

markers_leu_vs_ctrl_HSC <- FindMarkers(subset_cluster_HSC,
                                       ident.1 = "Leukemia",
                                       ident.2 = "Healthy",min.pct=0.15)

#Filter significant genes cluster HSC

markers_leu_vs_ctrl_HSC <- markers_leu_vs_ctrl_HSC %>%
  dplyr::filter(p_val_adj < 0.05) %>%
  dplyr::arrange(desc(avg_log2FC))

```
## Processing Prog cluster and find markers AML vs Control
```{r, echo=TRUE, results="hide", message=FALSE, warning=FALSE}
subset_cluster_Prog <- subset(seurat_obj, idents = "Prog")
table(subset_cluster_Prog@meta.data$group)
Idents(subset_cluster_Prog) <- "group"

#Find markers Prog
markers_leu_vs_ctrl_Prog <- FindMarkers(subset_cluster_Prog,
                                        ident.1 = "Leukemia",
                                        ident.2 = "Healthy",min.pct=0.15)
#Filter significant genes cluster Prog
 markers_leu_vs_ctrl_Prog <- markers_leu_vs_ctrl_Prog %>%
  dplyr::filter(p_val_adj < 0.05) %>%
  dplyr::arrange(desc(avg_log2FC))

```
## Processing HSC cluster and find markers postchemo vs prechemo
```{r, echo=TRUE, results="hide", message=FALSE, warning=FALSE}
HSC_condition_obj <- subset_cluster_HSC
table(Idents(HSC_condition_obj))
Idents(HSC_condition_obj) <- "Condition"

#Find markers HSC postchemo vs prechemo
markers_HSC_postchemo_vs_prechemo <- FindMarkers(HSC_condition_obj, ident.1 = "post_chemo", ident.2 = "pre_chemo", min.pct=0.15)
length(which(markers_HSC_postchemo_vs_prechemo$p_val_adj<0.05))

#Filter significant genes cluster HSC postchemo vs prechemo
markers_HSC_postchemo_vs_prechemo<- markers_HSC_postchemo_vs_prechemo%>%
  dplyr::filter(p_val_adj < 0.05) %>%
  dplyr::arrange(desc(avg_log2FC))
```

## Processing Prog cluster and find markers postchemo vs prechemo
```{r, echo=TRUE, results="hide", message=FALSE, warning=FALSE}
Prog_condition_obj <- subset_cluster_Prog
table(Idents(Prog_condition_obj))
Idents(Prog_condition_obj) <- "Condition"

#Find markers Pog postchemo vs prechemo
markers_Prog_postchemo_vs_prechemo <- FindMarkers(Prog_condition_obj, ident.1 = "post_chemo", ident.2 = "pre_chemo",min.pct=0.15)
length(which(markers_Prog_postchemo_vs_prechemo$p_val_adj<0.05))

#Filter significant genes cluster Prog postchemo vs prechemo
markers_Prog_postchemo_vs_prechemo<- markers_Prog_postchemo_vs_prechemo%>%
  dplyr::filter(p_val_adj < 0.05) %>%
  dplyr::arrange(desc(avg_log2FC))

```
##Save DE results
```{r echo=TRUE, results="hide", message=FALSE, warning=FALSE}
write.csv(markers_leu_vs_ctrl_HSC, "markers_leu_vs_ctrl_HSC.csv")
write.csv(markers_leu_vs_ctrl_Prog, "markers_leu_vs_ctrl_Prog.csv")
write.csv(markers_HSC_postchemo_vs_prechemo, "markers_HSC_postchemo_vs_prechemo.csv")
write.csv(markers_Prog_postchemo_vs_prechemo, "markers_Prog_postchemo_vs_prechemo.csv")

```
## Volcano plots

## Plot "Prog cluster, AML vs Control"
```{r, echo=TRUE, results="hide", message=FALSE, warning=FALSE}
top_leu_vs_ctrl_Prog <- markers_leu_vs_ctrl_Prog 

genes_Prog_leu_vs_ctrl_indexes <- c(1:10, (nrow(top_leu_vs_ctrl_Prog)-10):nrow(top_leu_vs_ctrl_Prog))
genes_Prog_leu_vs_ctrl<- rownames(top_leu_vs_ctrl_Prog[genes_Prog_leu_vs_ctrl_indexes, ])

top_leu_vs_ctrl_Prog <- top_leu_vs_ctrl_Prog |>
  mutate(gene_name = row.names(top_leu_vs_ctrl_Prog), 
         color = ifelse(top_leu_vs_ctrl_Prog$avg_log2FC<(-1)&top_leu_vs_ctrl_Prog$p_val_adj<0.05, 
                        "blue", ifelse(top_leu_vs_ctrl_Prog$avg_log2FC>1&top_leu_vs_ctrl_Prog$p_val_adj<0.05, 
                                       "red", 
                                       "grey")))

#Labeling genes
label_genes_prog <-top_leu_vs_ctrl_Prog[top_leu_vs_ctrl_Prog$gene_name %in% genes_Prog_leu_vs_ctrl, ]
#Plot
p_leu_vs_ctrl_Prog <- ggplot(top_leu_vs_ctrl_Prog, aes(x = avg_log2FC, y = -log10(p_val_adj))) +
  geom_point(aes(color = color), alpha = 0.7) +
  geom_text_repel(data = label_genes_prog,
                  aes(label = gene_name),
                  size = 5,
                  max.overlaps = 100,
                  segment.color = NA) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black") +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "black") +
  theme(
    plot.title = element_text(size = 20),
    axis.title.x = element_text(size = 18),
    axis.title.y = element_text(size = 18)) +
  labs(
    title = "Prog cluster, AML vs Control",
    x = "log2 Fold Change",
    y = "-log10 Adjusted p-value"
  ) +
  scale_color_identity()
p_leu_vs_ctrl_Prog 
#Save plot
ggsave("p_leu_vs_ctrl_Prog.pdf", plot = p_leu_vs_ctrl_Prog, width = 10, height = 8)
ggsave("p_leu_vs_ctrl_Prog.jpeg", plot = p_leu_vs_ctrl_Prog, width = 10, height = 8)
```
## Plot "HSC cluster, AML vs Control"
```{r, echo=TRUE, results="hide", message=FALSE, warning=FALSE}
top_leu_vs_ctrl_HSC <- markers_leu_vs_ctrl_HSC

genes_HSC_leu_vs_ctrl_indexes <- c(1:10, (nrow(top_leu_vs_ctrl_HSC)-10):nrow(top_leu_vs_ctrl_HSC))
genes_HSC_leu_vs_ctrl<- rownames(top_leu_vs_ctrl_HSC[genes_HSC_leu_vs_ctrl_indexes, ])

top_leu_vs_ctrl_HSC <- top_leu_vs_ctrl_HSC |>
  mutate(gene_name = row.names(top_leu_vs_ctrl_HSC), 
         color = ifelse(top_leu_vs_ctrl_HSC$avg_log2FC<(-1)&top_leu_vs_ctrl_HSC$p_val_adj<0.05,
                        "blue", ifelse(top_leu_vs_ctrl_HSC$avg_log2FC>1&top_leu_vs_ctrl_HSC$p_val_adj<0.05,
                                       "red",
                                       "grey")))
#Labeling genes
label_genes_HSC <- top_leu_vs_ctrl_HSC[top_leu_vs_ctrl_HSC$gene_name %in% genes_HSC_leu_vs_ctrl, ]

#Plot
p_leu_vs_ctrl_HSC <- ggplot(top_leu_vs_ctrl_HSC, aes(x = avg_log2FC, y = -log10(p_val_adj))) +
  geom_point(aes(color = color), alpha = 0.7) +
  geom_text_repel(data = label_genes_HSC,
                  aes(label = gene_name),
                  size = 6,
                  max.overlaps = 100,
                  segment.color = NA
  ) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black") +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "black") +
  theme(
    plot.title = element_text(size = 20),
    axis.title.x = element_text(size = 18),
    axis.title.y = element_text(size = 18)) +
  labs(
    title = "HSC cluster, AML vs Control",
    x = "log2 Fold Change",
    y = "-log10 Adjusted p-value"
  ) +
  scale_color_identity()
p_leu_vs_ctrl_HSC

#Save plot
ggsave("p_leu_vs_ctrl_HSC.pdf", plot = p_leu_vs_ctrl_HSC, width = 10, height = 8)
ggsave("p_leu_vs_ctrl_HSC.jpeg", plot = p_leu_vs_ctrl_HSC, width = 10, height = 8)
```
## Plot "HSC cluster, AML post-chemo vs pre-chemo"
```{r, echo=TRUE, results="hide", message=FALSE, warning=FALSE}
top_HSC_postchemo_vs_prechemo <- markers_HSC_postchemo_vs_prechemo 

genes_HSC_postchemo_vs_prechemo_indexes <- c(1:3, (nrow(top_HSC_postchemo_vs_prechemo)-3):nrow(top_HSC_postchemo_vs_prechemo))
genes_HSC_postchemo_vs_prechemo <- rownames(top_HSC_postchemo_vs_prechemo[genes_HSC_postchemo_vs_prechemo_indexes, ])

top_HSC_postchemo_vs_prechemo <- top_HSC_postchemo_vs_prechemo |>
  mutate(gene_name = row.names(top_HSC_postchemo_vs_prechemo), 
         color = ifelse(top_HSC_postchemo_vs_prechemo$avg_log2FC<(-1)&top_HSC_postchemo_vs_prechemo$p_val_adj<0.05,
                        "blue", ifelse(top_HSC_postchemo_vs_prechemo$avg_log2FC>1&top_HSC_postchemo_vs_prechemo$p_val_adj<0.05,
                                       "red",
                                       "grey")))
#Labeling genes
label_genes_HSC_post_vs_pre <- top_HSC_postchemo_vs_prechemo[top_HSC_postchemo_vs_prechemo$gene_name %in% genes_HSC_postchemo_vs_prechemo, ]

#Plot
p_leu_postchemo_vs_prechemo_HSC <- ggplot(top_HSC_postchemo_vs_prechemo, aes(x = avg_log2FC, y = -log10(p_val_adj))) +
  geom_point(aes(color = color), alpha = 0.7) +
  geom_text_repel(data = label_genes_HSC_post_vs_pre,
                  aes(label = gene_name),
                  size = 5,
                  max.overlaps = 100,
                  segment.color = NA) +
  
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black") +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "black") +
  theme(
    plot.title = element_text(size = 20),
    axis.title.x = element_text(size = 18),
    axis.title.y = element_text(size = 18)) +
  labs(
    title = "HSC cluster, AML post-chemo vs pre-chemo",
    x = "log2 Fold Change",
    y = "-log10 Adjusted p-value"
  ) +
  scale_color_identity()

p_leu_postchemo_vs_prechemo_HSC

#Save plot
ggsave("p_leu_postchemo_vs_prechemo_HSC.pdf", plot = p_leu_postchemo_vs_prechemo_HSC, width = 10, height = 8)
ggsave("p_leu_postchemo_vs_prechemo_HSC.jpeg", plot =p_leu_postchemo_vs_prechemo_HSC, width = 10, height = 8)

```



## Plot "Prog cluster, AML post-chemo vs pre-chemo"
```{r, echo=TRUE, results="hide", message=FALSE, warning=FALSE}
top_Prog_postchemo_vs_prechemo <-markers_Prog_postchemo_vs_prechemo 

genes_Prog_postchemo_vs_prechemo_indexes <- c(1:10, (nrow(top_Prog_postchemo_vs_prechemo)-10):nrow(top_Prog_postchemo_vs_prechemo))
genes_Prog_postchemo_vs_prechemo<- rownames(top_Prog_postchemo_vs_prechemo[genes_Prog_postchemo_vs_prechemo_indexes, ])

top_Prog_postchemo_vs_prechemo <- top_Prog_postchemo_vs_prechemo |>
  mutate(gene_name = row.names(top_Prog_postchemo_vs_prechemo), 
         color = ifelse(top_Prog_postchemo_vs_prechemo$avg_log2FC<(-1)&top_Prog_postchemo_vs_prechemo$p_val_adj<0.05,
                        "blue", ifelse(top_Prog_postchemo_vs_prechemo$avg_log2FC>1&top_Prog_postchemo_vs_prechemo$p_val_adj<0.05,
                                       "red",
                                       "grey")))
#Labeling genes
label_genes_Prog_post_vs_pre <- top_Prog_postchemo_vs_prechemo[top_Prog_postchemo_vs_prechemo$gene_name %in% genes_Prog_postchemo_vs_prechemo, ]

#Plot
p_leu_postchemo_vs_prechemo_Prog <- ggplot(top_Prog_postchemo_vs_prechemo, aes(x = avg_log2FC, y = -log10(p_val_adj))) +
  geom_point(aes(color = color), alpha = 0.6) +
  geom_text_repel(data = label_genes_Prog_post_vs_pre,
                  aes(label = gene_name),
                  size = 5,
                  max.overlaps = 100,
                  segment.color = NA
  ) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black") +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "black") +
  theme(
    plot.title = element_text(size = 20),
    axis.title.x = element_text(size = 18),
    axis.title.y = element_text(size = 18))+
  labs(
    title = "Prog cluster, AML post-chemo vs pre-chemo",
    x = "log2 Fold Change",
    y = "-log10 Adjusted p-value"
  ) +
  scale_color_identity()
p_leu_postchemo_vs_prechemo_Prog

#Save plot
ggsave("p_leu_postchemo_vs_prechemo_Prog.pdf", plot = p_leu_postchemo_vs_prechemo_Prog, width = 10, height = 8)
ggsave("p_leu_postchemo_vs_prechemo_Prog.jpeg", plot =p_leu_postchemo_vs_prechemo_Prog, width = 10, height = 8)
```
## Overrepresentation_enrichR
##Import dbs
```{r, echo=TRUE, results="hide", message=FALSE, warning=FALSE}
#dbs <- listEnrichrDbs()
dbs <- c("GO_Biological_Process_2025", "KEGG_2021_Human", "MSigDB_Hallmark_2020")

```
##function - enrichR enrichment
```{r, echo=TRUE, results="hide", message=FALSE, warning=FALSE}
#function - enrichR enrichment, 
  #arguments:de_table,name_prefix, sample_names = c("Control", "Tumor"), cluster
r_enrichment <- function(de_table, name_prefix, output_dir = "enrichr_results", 
                         plot_title = NULL, sample_names = NULL, cluster = NULL, db_names = NULL) {
  
   if (!dir.exists(output_dir)) dir.create(output_dir)
  
  up_genes <- rownames(de_table[de_table$p_val_adj < 0.05 & de_table$avg_log2FC > 1, ])
  down_genes <- rownames(de_table[de_table$p_val_adj < 0.05 & de_table$avg_log2FC < -1, ])
  all_genes <- rownames(de_table[de_table$p_val_adj < 0.05, ])
  
# enrichR
  up_enrich <- enrichr(up_genes, dbs)
  down_enrich <- enrichr(down_genes, dbs)
  all_enrich <- enrichr(all_genes, dbs)

  # plot title
  if (is.null(plot_title)) {
    
      db_labels <- c(
      "GO_Biological_Process_2025" = "GO Biological Process",
      "KEGG_2021_Human" = "KEGG",
      "MSigDB_Hallmark_2020" = "MSigDB"
    )
db_title <- ifelse(is.null(db_names), "GO Biological Process", db_labels[db_names])
    
    plot_title <- paste(db_title, "enrichment analysis,",
                        ifelse(!is.null(sample_names) & length(sample_names) == 2,
                               paste(sample_names[1], "vs", sample_names[2]),
                               "Samples Comparison"),
                        ifelse(!is.null(cluster), paste("cluster", cluster), ""))
  }
  
  saved_plots <- list()
  
  for (direction in c("up", "down", "all")) {
    enrich_list <- switch(direction,
                          up = up_enrich,
                          down = down_enrich,
                          all = all_enrich)
    
    gene_set <- switch(direction,
                       up = up_genes,
                       down = down_genes,
                       all = all_genes)
    
    for (db in names(enrich_list)) {
      df <- enrich_list[[db]]
      
      print(paste("Direction:", direction))
      print(paste("DB:", db))
      print(paste("Genes:", length(gene_set)))
      print(head(gene_set))
      print(paste("Rows in df:", if (!is.null(df)) nrow(df) else "NULL"))
      
      db_title <- db_labels[db]
       if (!is.null(df) && nrow(df) > 0) {
        write.csv(df, file = file.path(output_dir, paste0(name_prefix, "_", direction, "_", db, ".csv")), row.names = FALSE)
        
        # Overlap to numeric
        df <- df %>%
          mutate(Overlap = as.character(Overlap),
                 GeneCountRatio = sapply(Overlap, function(x) {
                   nums <- as.numeric(strsplit(x, "/")[[1]])
                   if (length(nums) == 2) return(nums[1] / nums[2])
                   else return(NA)  
                 })) %>%
          mutate(term = factor(Term, levels = rev(Term)),
                 negLogP = -log10(Adjusted.P.value))
        
        top_terms <- df %>%
          arrange(Adjusted.P.value) %>%
          head(25)
        
        p <- ggplot(top_terms, aes(x = negLogP, y = term, color = Adjusted.P.value, size = GeneCountRatio)) +
          geom_point() +
          scale_color_gradient(low = "blue", high = "red") +
          theme_minimal() +
          labs(
            title = paste(db_title, "enrichment analysis,", 
                          ifelse(!is.null(sample_names) & length(sample_names) == 2,
                                 paste(sample_names[1], "vs", sample_names[2]),
                                 "Samples Comparison"),
                          ifelse(!is.null(cluster), paste("cluster", cluster), "")),  
            x = "-log10 Adjusted p-value",
            y = "Term",
            color = "adj. p-value",
            size = "Gene Count Ratio"
          ) +
          theme(
            plot.title = element_text(size = 16, face = "bold"),  
            axis.title = element_text(size = 14),  
            axis.text = element_text(size = 14),  
            legend.title = element_text(size = 12),  
            legend.text = element_text(size = 12)  
          )
        
        saved_plots[[paste0(direction, "_", db)]] <- p
        #different size
        if (db == "GO_Biological_Process_2025") {
          ggsave(filename = file.path(output_dir, paste0(name_prefix, "_", direction, "_", db, "_dotplot.pdf")), plot = p, width = 17, height = 7)
          ggsave(filename = file.path(output_dir, paste0(name_prefix, "_", direction, "_", db, "_dotplot.jpeg")), plot = p, width = 17, height = 7)
        } else {
          ggsave(filename = file.path(output_dir, paste0(name_prefix, "_", direction, "_", db, "_dotplot.pdf")), plot = p, width = 12, height = 7)
          ggsave(filename = file.path(output_dir, paste0(name_prefix, "_", direction, "_", db, "_dotplot.jpeg")), plot = p, width = 12, height = 7)
        }
      }
    }
  }
  
    return(list(
    up = up_enrich,
    down = down_enrich,
    all = all_enrich,
    saved_plots = saved_plots  
  ))
}

```
## enrichR cluster Prog
```{r, echo=TRUE, results="hide", message=FALSE, warning=FALSE}
enrich_leu_vs_ctrl_Prog <- r_enrichment(de_table = markers_leu_vs_ctrl_Prog,
                                        name_prefix = "leu_vs_ctrl_Prog", 
                                        sample_names = c("AML", "Control"),
                                       cluster = "Prog")
```
## enrichR cluster HSC
```{r, echo=TRUE, results="hide", message=FALSE, warning=FALSE}
enrich_leu_vs_ctrl_HSC <- r_enrichment(de_table = markers_leu_vs_ctrl_HSC,
                                        name_prefix = "leu_vs_ctrl_HSC", 
                                        sample_names = c("AML", "Control"),
                                        cluster = "HSC")
```
## enrichR cluster Prog postchemo_vs_prechemo
```{r, echo=TRUE, results="hide", message=FALSE, warning=FALSE}
enrich_postchemo_vs_prechemo_Prog <- r_enrichment(de_table = markers_Prog_postchemo_vs_prechemo,
                                       name_prefix = "postchemo_vs_prechemo_Prog", 
                                       sample_names = c("Post-chemo", "Pre-chemo"),
                                       cluster = "Prog")

```
## enrichR cluster HSC postchemo_vs_prechemo
```{r, echo=TRUE, results="hide", message=FALSE, warning=FALSE}
enrich_postchemo_vs_prechemo_HSC <- r_enrichment(de_table = markers_HSC_postchemo_vs_prechemo,name_prefix = "postchemo_vs_prechemo_HSC", sample_names = c("Post-chemo", "Pre-chemo"),cluster = "HSC") #no terms
```
## Display the saved plots for cluster Prog postchemo vs prechemo
```{r, echo=TRUE, results="hide", message=FALSE, warning=FALSE}

for (plot_name in names(enrich_postchemo_vs_prechemo_Prog$saved_plots)) {
print(enrich_postchemo_vs_prechemo_Prog$saved_plots[[plot_name]])
}

```

## Display the saved plots for cluster Prog AML vs Control
```{r, echo=TRUE, results="hide", message=FALSE, warning=FALSE}

for (plot_name in names(enrich_leu_vs_ctrl_Prog$saved_plots)) {
print(enrich_leu_vs_ctrl_Prog$saved_plots[[plot_name]])
}

```
## Display the saved plots for cluster HSC AML vs Control
```{r, echo=TRUE, results="hide", message=FALSE, warning=FALSE}

for (plot_name in names(enrich_leu_vs_ctrl_HSC$saved_plots)) {
print(enrich_leu_vs_ctrl_HSC$saved_plots[[plot_name]])
}

```